---
title: "Sexual Health and non-English Language in Chicago, Illinois"
output:
  html_document: github_document
  pdf_document: default
  github_document: default
---
Nathan Eckert ne4256

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
  The data included in this exploratory analysis are from the official City of Chicago website. The first data set called Languages Spoken in Chicago was collected from 2008 to 2012. It includes data about the predominant non-English language spoken in each Chicago community. It also includes the percent of residents over five, who speak English "less than very well" in addition to the counts of the spoken languages. The data was compiled from U.S. Census Bureau's 2008-2012 American Community Survey 5-year estimates. The second data set is named Public Heath Statistics from 2005 to 2011 and includes various health indicators organized by community area of Chicago. It was collected from multiple sources including census data, Chicago Dept. of Public Health, and reports accessible through local infectious disease laws.

  I wanted to investigate these data sets as I was curious if groups of people who spoke little English, or specific ethnicities were more prone to specific conditions.
I expect that areas with lower education and less skill with the English language will have more diseases such as gonorrhea. I also assume that rates of STDs will be correlated with high rates of pregnancy, suggesting lack of contraceptive use.

*Source URLs*
https://data.cityofchicago.org/Health-Human-Services/Census-Data-Languages-spoken-in-Chicago-2008-2012/a2fk-ec6q
https://data.cityofchicago.org/Health-Human-Services/Public-Health-Statistics-Selected-public-health-in/iqnk-2tcu


```{r}
#Opened data sets and renamed to simpler names.

library(readxl)
Chicago_Health_short_ <- read_excel("Chicago Health (short).xlsx")
Chicago_Languages_short_ <- read_excel("Chicago Languages (short).xlsx")

chic_lang <- Chicago_Languages_short_
chic_health <- Chicago_Health_short_
```


## Tidy

```{r}

library(tidyverse)
library(dplyr)

head(chic_lang)
#separate Non-English Lang (%) into two separate columns for predominant non-English language spoken and percent that speak English less than very well.
chic_lang <- chic_lang%>%
  separate("NON-ENG LANG (%)", into = c("NON-ENGLISH LANG", "Percent under well"))
head(chic_lang)

#This code turns all spaces into periods and all commas into spaces.

names(chic_health)<-str_replace_all(names(chic_health), c(" " = "." , "," = "" ))
names(chic_lang)<-str_replace_all(names(chic_lang), c(" " = "." , "," = "" ))

#This changes any missing values like "." into NAs.
chic_health <- na_if(chic_health, ".")
chic_lang <- na_if(chic_lang, ".")

```
The data sets are now tidy and ready to be joined.

## Join/ Merge


```{r}
#I will join the two data sets chic_lang and chic_health by community name and remove a redundant column "community.y".
chicago <- left_join(chic_health, chic_lang, by = "Community.Name")%>%
  select(!"Community.y")

#Fixed percent and gonorrhea.m being chr, now is numeric.
chicago <- chicago%>%
  mutate_at(vars(`Percent.under.well`),as.numeric)%>%
  mutate_at(vars(`Gonorrhea.M`),as.numeric)
chicago
```
I used left join with chic_health first because chic_lang had an unpaired observation for the city of Chicago as a whole not a community. This is not what I am interested in investigating so I used left join to remove it. There were no other removals of observations. A redundant column was removed.
The output is the joined data set called "chicago"

## Summary Statistics

```{r}
#removed health variables that were not of interest/related to sexual health and poverty
chicago_sex <- chicago%>%
  select(-(9:14),-(16:18),-23,-(25:26))
chicago_sex
```
Chicago_sex now contains data I am interested in. This is largely sexual health information.

```{r}
#Remove rows where there is an NA for Gonorrhea.
chicago_sex.gon<- chicago_sex%>%
  filter(!is.na(Gonorrhea.F),!is.na(Gonorrhea.M))

  #Find average Gonorrhea rate for females with below code.
#summarize(mean(Gonorrhea.F))

#Create proportion for female Gonorrhea rates above average.
gon.data <- chicago_sex.gon%>%
  mutate(gon.rate = Gonorrhea.F/906.0516)%>%
  #put greatest above average rate first
  arrange(desc(gon.rate))%>%
  select(c(gon.rate, 'Percent.under.well', Community.x))%>%
  #remove some points where percent.under.well was unclear
  filter(!Community.x == 47)%>%
 filter(!Community.x == 45)

gon.data
```
This created gon.data which gives a quick look at the relationship between gon.rate (ratio of gonorrhea rates per 100,000 females ages 15-44) and Percent.under.well (percent of community which speaks English less than very well). The results appear to be that more gonorrhea is associated with knowing more English. 

```{r}

chicago_sex
#grouped by predominant second non-English language and checked various averages
chicago_sex%>%
 filter(!Community.x == 45)%>%
  group_by(`NON-ENGLISH.LANG`)%>%
  summarize(mean(Birth.Rate),mean(Below.Poverty.Level),mean(No.HS.Diploma),mean(Infant.Mort.Rate),mean(Percent.under.well))



```
This gave an overview of some averages of interest grouped by Predominant non-English Language. Spanish speakers appear to have highest birth rate. African speakers appear to have highest infant mortality rate. Polish appear to have the lowest numbers below poverty level.


```{r}
#Find count of each community by Predominant non_English Language spoken.
chicago_sex%>%
  group_by(`NON-ENGLISH.LANG`)%>%
  count()
```
Shows disproportionate sample sizes. Majority is Spanish speakers.
 
 **Creating various summary stats for Gonorrhea.F**
```{r}
#strong correlation with male gonorrhea
cor(chicago_sex.gon$Gonorrhea.F,chicago_sex.gon$Gonorrhea.M)

#five number summary for gonorhea rate for chicago (min, lower IQR, median, upper IQR, max)
fivenum(chicago_sex.gon$Gonorrhea.F)
#huge range of rates

sd(chicago_sex.gon$Gonorrhea.F)
#standard deviation is very wide, agreeing with wide range

#distinct number of female gonorrhea rates
n_distinct(chicago_sex.gon$Gonorrhea.F)
```
 
 
**Gonorrhea Statistics Summary**

 min  | lower IQR| median| upper IQR| max      | sd   | n   |cor with males
------|----------|-------|----------|----------|------|-----|--------------
 50.3 |  147.8   |569.6  |1583.7    |    3193.3| 896.9| 64  |0.93


Of note is a high variation and strong correlation with male gonorrhea rates.

```{r}
#code used earlier that groups by categorical non-English Predominant Language 
chicago_sex%>%
 filter(!Community.x == 45)%>%
  group_by(`NON-ENGLISH.LANG`)%>%
  summarize(mean(Birth.Rate),mean(Gonorrhea.F) ,mean(No.HS.Diploma),mean(Infant.Mort.Rate),mean(Percent.under.well))

#Gives median and IQR of gonorrhea by pred non-English language
chicago_sex%>%
  group_by(`NON-ENGLISH.LANG`)%>%
  summarize(median(Gonorrhea.F, na.rm=T),IQR(Gonorrhea.F, na.rm=T))

#used to check Polish observations of gonorrhea 
chicago_sex%>%
  filter(chicago_sex$`NON-ENGLISH.LANG` == 'POLISH')
```
**Female Gonorrhea median and IQR by predominant non-English Language**

| Pred. non-English lang.  | median Gonorrhea.F| IQR Gonorrhea.F|  
|-----------------------|-------------------|----------------|
| Chinese                | 220              | 130            | |-----------------------|-------------------|----------------|
|African                | 2124              |  984           |           |
|-----------------------|-------------------|----------------|
|Spanish                | 610               |   1514         |

Spanish speakers have greatest IQR. This could be due to larger sample size.
Note: Polish not included as no female gonorrhea observations were present.The single male rate was 70 (per 100,000 ages 15-44).


## Visualizations

```{r}
#correlation heat map
#choosing variables to correlate
chicago_sex.short<- chicago_sex%>%
  select(c(Birth.Rate, Infant.Mort.Rate, Gonorrhea.F, Percent.under.well, No.HS.Diploma, Teen.Birth.Rate))


# Build a correlation matrix between all numeric variables
chicago_sex.short_num <- chicago_sex.short %>%
  select_if(is.numeric) 
cor(chicago_sex.short_num, use = "pairwise.complete.obs")

# Making a correlation heatmap with geom_tile
cor(chicago_sex.short_num, use = "pairwise.complete.obs") %>%
  # Save as a data frame
  as.data.frame %>%
  # Convert row names to an explicit variable
  rownames_to_column %>%
  # Pivot so that all correlations appear in the same column
  pivot_longer(-1, names_to = "other_var", values_to = "correlation") %>%
  ggplot(aes(rowname, other_var, fill=correlation,)) +
  # Heatmap with geom_tile
  geom_tile() +
  # Change the scale to make the middle appear neutral
  scale_fill_gradient2(low="blue",mid="white",high="orange") +
  # Overlay values
  geom_text(aes(label = round(correlation,2)), color = "black", size = 4) +
  # Give title and labels
  labs(title = "Correlation matrix for the dataset chicago_sex.short", x = "variable 1", y = "variable 2")+
  theme(axis.text.x = element_text(angle=90))

```

Strongest correlation is between the percent who speak English less than very well and not having a high school diploma. This may be due to growing up outside of the US. Another strong correlation (.75) is between Teen Birth Rate and Gonnorhea rates in females. This may suggest a lack of contraception use. Notably there is little to no correlation between gonorrhea rates and not having a high school diploma. This may suggest high school sexual education is either absent or ineffective in this regard.


```{r}
#plot of Gonorrhea, Infant mortality, and Prenatal Care in 1st trimester

library(ggplot2)

#make variable for >70.8 (above us avg)
chicago_sex
chicago_sex.gon
chicago_sex.gon%>%
  mutate(first.trim = case_when(Prenatal.Care.First.Trim>70.8 ~ "above",
                                Prenatal.Care.First.Trim<=70.8 ~ "below"))%>%
  ggplot(aes(x = Gonorrhea.F, y = Infant.Mort.Rate, color = first.trim, stat = "summary")) + geom_point() +
  #custom colors
  scale_color_brewer(palette="Dark2")+
  #titles and line to make more clear
  ggtitle("Gonorrhea and Infant Mortality by Prenatal Care in First Trimester") +
  ylab("Infant Mortality (per 1000 live births)")+ xlab("Female Gonorrhea (per 100,000 ages 15-44)")+
  geom_smooth(method = "lm") + labs(color = "US average % Prenatal care \n in first trimester (70.8)")

```

This plot shows that there appears to be a positive relationship between infant mortality and female gonorrhea rates. This is expected as STDs can complicate pregnancy and may serve as proxies for other health issues. The graph shows the difference between those communities that seek prenatal care in the first trimester in percentages above and below the US average for 2009. It appears those communities which received care at higher rates did had lower infant mortality than those who did. That is seeking early care led to worse outcomes for infants. This seems unlikely and may be either an insignificant difference or a result of sample sizes.

```{r}
#make variable for >70.8 (above us avg)
chicago_sex
chicago_sex.gon
chicago_sex.gon%>%
  mutate(first.trim = case_when(Prenatal.Care.First.Trim>70.8 ~ "above",
                                Prenatal.Care.First.Trim<=70.8 ~ "below"))%>%
  ggplot(aes(x = `NON-ENGLISH.LANG`, fill= first.trim)) +
  #separate graphs and make height equal mean Infant.Mort.Rate
  geom_bar(aes(y = Infant.Mort.Rate), stat = "summary", fun="mean", position = "dodge")+ 
 #custom colors
  scale_fill_brewer(palette="Dark2")+
  #titles and line to make more clear
  ggtitle("Infant Mortality and Predominant non-English Language \nby Prenatal Care in First Trimester") +
  ylab("Infant Mortality (per 1000 live births)")+ xlab("Predominant non-English Language")+ labs(fill="US average % Prenatal care \n in first trimester (70.8)")
  
#table to confirm means
chicago_sex.gon%>%
  group_by(`NON-ENGLISH.LANG`)%>%
  summarize(mean(Infant.Mort.Rate))


```
This plot was created to assess if particular language speakers had higher Infant Mortality Rates regardless of when they received prenatal care. If there were irregularities in this regard, a possible suggestion could be certain ethnic groups received worse care at Chicago hospitals. The data agrees with the previous plot's suggestion that early prenatal care is associated with higher infant mortality. This remains perplexing. The plot shows a difference in health outcomes for infants to be similar across language speakers suggesting each speaker received comparable care. No Chinese speakers received prenatal care in percentages less than the US average.

```{r}
chicago_sex.gon%>%
  arrange(`NON-ENGLISH.LANG`)

chicago_sex%>%
  ggplot(aes(y = Birth.Rate, x = No.HS.Diploma, color = `NON-ENGLISH.LANG`)) + geom_point() +
  #custom colors
  scale_color_brewer(palette="Accent")+
  #titles and line to make more clear
  ggtitle("No High School degree and Birth Rate by Predominant non-English Language") +
  ylab("Birth Rate (per 1000 persons)")+ xlab("No High School Diploma (% over 25 years old)")+
  labs(color = "Predominant non-English Language")+
  #add regression lines without error band (they covered data)
  geom_smooth(method = "lm", se = FALSE)

```

This plot compares Birth Rate to rate of not having a high school diploma. This illustrates my lack of non-Spanish speaking communities. The expected trend is that Birth Rate increases as No High School Diploma does. The relationships are not clear in all but Spanish speakers. Polish is slightly positive, but African only has two data points, and Chinese may have an outlier. It is unclear if the relationship between Birth Rate and No High School Diploma is influenced by Predominant non-English Language spoken.



## Clustering

```{r}
library(cluster)
#clustering with two variables: Birth.Rate and Below.Poverty.Level
myclus_data<- chicago_sex%>%
  select(Birth.Rate, Below.Poverty.Level)
myclus_data

#perform pam
mypam1 <- myclus_data %>%
  pam(k=3)
mypam1

#make clusters a part of data
mypamclus <- myclus_data %>%
  mutate(cluster = as.factor(mypam1$clustering))
mypamclus

#graph
mypamclus %>% 
  ggplot(aes(Below.Poverty.Level, Birth.Rate, color = cluster)) +
  geom_point()+ ggtitle("Clusters of Below Poverty Level and Birth Rate")+
  xlab("Percent below Poverty Level")+ ylab("Birth Rate")+
  labs(color = "Cluster")+
  scale_color_brewer(palette="Dark2")

#Make graph data include NON-ENGLISH.LANG variable.
graph <- mypamclus %>% 
  mutate("NON-ENGLISH.LANG" = chicago_sex$"NON-ENGLISH.LANG")
graph

#Graph comparing clusters to predominant non-English language.
graph%>%
  ggplot(aes(y = Birth.Rate, x = Below.Poverty.Level, shape = `NON-ENGLISH.LANG`, color = cluster)) + geom_point()+xlab("Percent below Poverty Level")+ ylab("Birth Rate")+
  ggtitle("Clusters and Predominant non-English Language of Percent Below Poverty \nLevel and Birth Rate")+
  labs(shape = "Predominant non-English \nLanguage", color = "Cluster")+
  scale_color_brewer(palette="Dark2")

```

Included are two graphs of the clusters between Birth Rate and Percent Below Poverty level. One includes the labels of Predominant non-English Language spoken. The data seems to show that there are clusters largely depending on Percent Below Poverty Level. It is difficult to say if the predominant language can predict cluster. That is, the data is not naturally separated by predominant language. Overall it seems that more poverty leads to higher birth rates which can be due to a number of factors. Just by observation it appears that Chinese speakers have a lower Birth Rate. Polish speakers tend to have the lowest Percent Below Poverty Level. Spanish speakers seem evenly distributed and African speakers are too few to make reasonable comments about.

Check number of clusters
```{r}
# Number of clusters based on WSS (within sum of squares, i.e. tightness of clusters)
# create an empty vector
wss <- vector()

# Find WSS for different values of k using a loop
for(i in 1:10){
temp <- chicago_sex %>%
  select(Birth.Rate,Below.Poverty.Level) %>%
  kmeans(.,i)
wss[i]<-temp$tot.withinss
}

# Represent the value of WSS depending on the number of clusters
ggplot() + 
  geom_point(aes(x=1:10,y=wss)) +
  geom_path(aes(x=1:10,y=wss)) +
  xlab("clusters") +
  scale_x_continuous(breaks=1:10)+
  ggtitle("WSS and Number of Clusters")

```

I think three clusters is best number to minimize WSS while keeping cluster number small. I will leave the cluster number as three.





